---
title: "MKTG776 HW7"
author: "Jordan Farrer"
date: '2017-03-29'
output:
  bookdown::pdf_document2:
    fig_caption: yes
    number_sections: yes
    toc: no
  bookdown::html_document2:
    css: style.css
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = 'center', size = 'small', fig.pos = 'H',
                      fig.width = 6, fig.height = 3.5)

knitr::knit_hooks$set(
          size = function(before, options, envir) {
                  if (before) return(paste0("\n \\", options$size, "\n\n"))
                  else return("\n\n \\normalsize \n")
                  }
          , inline = function(x) {if(!is.numeric(x)){ x }else{ prettyNum(round(x,2), big.mark=",") } }
          )
```

```{r report-load-packages, results = 'hide', include = FALSE}
pacman::p_load(tidyverse, forcats, pander, stringr, lubridate, ggrepel, extrafont, jrfTools)
```

```{r report-additional-setup, include = FALSE}
options(scipen=999)
panderOptions('digits', 8)
panderOptions('round', 4)
panderOptions('keep.trailing.zeros', TRUE)
```

# Results 

Below is the business travel dataset used for fitting latent-class count models (NBD and Poission):

```{r load_data, echo = FALSE}
biz_travel_data <- readxl::read_excel("Biz travel data.xlsx", sheet = 1, 
                                      col_names = c('N', 'observed'), skip = 2)
biz_travel_observed <- (readxl::read_excel("Biz travel data.xlsx") %>% names())[2] %>% as.integer()
```

```{r echo = FALSE}
biz_travel_data %>%
  pander(caption = paste0("Business Travel Data for ", biz_travel_observed, " People"))
```


```{r NBD Function, echo = FALSE}
# For Zero-inflated Negative Binomial Distribution, calculates P(X=x) formula
fn_zinbd_formula <- function(x, r, alpha, pi) {
  p_x <- exp(lgamma(r + x) - (lgamma(r) + lfactorial(x))) * (alpha / (alpha + 1))^r * (1 / (alpha + 1))^x
  if(x == 0) {
    return(pi + (1 - pi) * p_x)  
  } else {
    return((1 - pi) * p_x)  
  }
}

# Deals with X+ situation
fn_zinbd_px <- function(x, r, alpha, pi) {
  x1 <- as.integer(str_replace(x, "\\+" ,""))
  if (str_detect(x, "\\+")) {
    return(1-sum(purrr::map_dbl(0:(x1-1), fn_zinbd_formula, r, alpha, pi)))
  } else {
    return(fn_zinbd_formula(x1, r, alpha, pi))
  }
}

# Calculates the log-likelihood of the NBD (including zero-inflated)
fn_zinbd_ll <- function(par, data, zero_inflated) {
  
  pi <- if_else(zero_inflated, par[3], 0)
  
  data2 <- 
    data %>%
    rowwise() %>%
    mutate(p_x = fn_zinbd_px(x = N, r = par[1], alpha = par[2], pi = pi)) %>%
    mutate(ll = observed * log(p_x))
  
  return(-sum(data2$ll))
}


fn_zinbd_model <- function(model, data, zero_inflated) {
  
  init_par <- list(nbd = list(start = c(1,1), lower = c(0,0), upper = c(Inf,Inf)), 
                   zinbd = list(start = c(1,1,.5), lower = c(0,0,0), upper = c(Inf,Inf,1)))
  
  init_par2 <- init_par[[zero_inflated + 1]]
  
  pars <- nlminb(start = init_par2$start, fn_zinbd_ll, lower = init_par2$lower, 
                 upper = init_par2$upper, data = data, zero_inflated = zero_inflated)$par
  return(
    data_frame(model = model, r = pars[1], alpha = pars[2], pi = if_else(zero_inflated, pars[3], NA_real_))
  )
  
}
```

```{r nbd_params, echo = FALSE}
nbd_params <- 
fn_zinbd_model("NBD", biz_travel_data, FALSE) %>%
  bind_rows(
    fn_zinbd_model("ZI-NBD", biz_travel_data, TRUE)    
  ) 
```

Below are the parameter estimates for the NBD and zero-inflated NBD models:

```{r, echo = FALSE}
nbd_params %>%
  pander(caption = "Parameter Estimates for NBD and Zero-Inflated NBD Models", missing = "")
```

```{r Latent-Class Poission Functions, echo = FALSE}
# Poission with arbitary number of lambdas and thetas
fn_lcp_formula <- function(x, lambdas, thetas) {
  p_x <- sum(dpois(x, lambdas) * exp(thetas) / sum(exp(thetas)))
  return(p_x)
}

# Deals with X+ situation
fn_lcp_px <- function(x, lambdas, thetas) {
  x1 <- as.integer(str_replace(x, "\\+" ,""))
  if (str_detect(x, "\\+")) {
    return(1-sum(purrr::map_dbl(0:(x1-1), fn_lcp_formula, lambdas, thetas)))
  } else {
    return(fn_lcp_formula(x1, lambdas, thetas))
  }
}

# Calculates the log-likelihood of the NBD (including zero-inflated)
fn_lcp_ll <- function(start, data, seg) {
  
  lambdas <- start[1:seg]
  if (seg > 1) {
    thetas <- c(start[(seg + 1):length(start)], 0)
  } else { 
    thetas <- 0
  }
  
  data2 <- 
    data %>%
    rowwise() %>%
    mutate(p_x = fn_lcp_px(x = N, lambdas, thetas)) %>%
    mutate(ll = observed * log(p_x))
  
  return(-sum(data2$ll))
}


fn_lcp_model <- function(model, data, seg) {
  
  start <- c(runif(seg, 1, 1), runif(seg-1, -1, 1))
  lower <- c(rep(0.01, seg), rep(-Inf, seg - 1))
  upper <- rep(Inf, seg)
  
  pars <- nlminb(start = start, fn_lcp_ll, lower = lower, upper = upper,  
                 data = data, seg = seg, control = list(x.tol = 1e-10))$par
  
  
  if (seg > 1) {
    parameters <- c(rep("lambda", seg), rep("theta", seg))
    thetas <- exp(c(pars[(seg + 1):length(start)], 0))
    estimates <- c(pars[1:seg], thetas / sum(thetas))
  } else {
    parameters <- "lambda"
    estimates <- pars
  }

  return(
    data_frame(model = rep(model, length(estimates)), parameter = parameters, estimate = estimates)
  )
}
```

```{r lcp_params, echo=FALSE}
lcp_params <- 
  fn_lcp_model("1-Seg", biz_travel_data, 1) %>%
    bind_rows(
      fn_lcp_model("2-Seg", biz_travel_data, 2), 
      fn_lcp_model("3-Seg", biz_travel_data, 3),
      fn_lcp_model("4-Seg", biz_travel_data, 4)
    )
```

We fit 4 Poission models - one with 1 segments, one with 2 segments, etc. Below are the parameters estimates for lambda (mean of the segments) and theta (% of travelers in each segment).

```{r lcp_params2, echo = FALSE}
lcp_params2 <- 
  lcp_params %>%
    group_by(model, parameter) %>%
    mutate(num = row_number()) %>%
    spread(num, estimate) %>%
    rename(`Seg 1` = `1`,`Seg 2` = `2`,`Seg 3` = `3`,`Seg 4` = `4`) %>%
    ungroup() %>%
    arrange(model, parameter) 

lcp_params2  %>%
  pander(caption = "Parameter Estimates for Latent-Class Poisson Models", missing = "")
```


```{r NBD Results, echo = FALSE}
fn_zinbd_ll_results <- function(data, r, alpha, pi, total_obs) {
  
  params = sum(!is.na(c(r, alpha, pi)))
  pi <- ifelse(!is.na(pi), pi, 0.0)
  
  data2 <- 
    data %>%
    rowwise() %>%
    mutate(p_x = fn_zinbd_px(x = N, r = r, alpha = alpha, pi = pi)) %>%
    ungroup() %>%
    mutate(
      ll = observed * log(p_x)
      , expected =  total_obs * p_x
      , chisq = (observed - expected)^2 / expected
    ) %>%
    summarise(
          ll = sum(ll)
        , chisq = sum(chisq)
        , percent_expected = sum(expected > 5) / n()
        , cells = n()
      ) %>%
    mutate(
      BIC = -2 * ll + params * log(total_obs)
      , p.value = pchisq(chisq, df = cells - params -1, lower.tail = FALSE)
      , params = params
    )
  
  if (data2$percent_expected < 0.8) {
    stop("Less than 80% of the cells have more than 5 counts")
  }
  
  return(
    data2 %>%
      select(
        LL = ll
        , `# params` = params
        , BIC = BIC
        , `$\\chi^2\\,p-value$` = p.value
      )
  )
  
}

nbd_results <- 
  nbd_params %>%
    crossing(biz_travel_data) %>%
    group_by(model, r, alpha, pi) %>%
    nest(.key = travel_data) %>%
    mutate(pred = pmap(list(travel_data, r, alpha, pi), fn_zinbd_ll_results, biz_travel_observed)) %>%
    select(model, pred) %>%
    unnest()
```

```{r Poisson Results, echo = FALSE}
# Poission with arbitary number of lambdas and thetas
fn_lcp_formula2 <- function(x, lambdas, thetas) {
  if (length(lambdas) == 1) {
    p_x <- sum(dpois(x, lambdas))
  } else {
    p_x <- sum(dpois(x, lambdas) * thetas)
  }
  return(p_x)
}

# Deals with X+ situation
fn_lcp_px2 <- function(x, lambdas, thetas) {
  x1 <- as.integer(str_replace(x, "\\+" ,""))
  if (str_detect(x, "\\+")) {
    return(1-sum(purrr::map_dbl(0:(x1-1), fn_lcp_formula2, lambdas, thetas)))
  } else {
    return(fn_lcp_formula2(x1, lambdas, thetas))
  }
}

fn_lcp_ll_results <- function(data, lambdas, thetas, total_obs) {
  
  if (length(lambdas) == 1) {
    params <- 1
  } else {
    params <- sum(!is.na(c(lambdas, thetas))) - 1
  }
    
  data2 <- 
    data %>%
    rowwise() %>%
    mutate(p_x = fn_lcp_px2(x = N, lambdas, thetas)) %>%
    ungroup() %>%
    mutate(
      ll = observed * log(p_x)
      , expected =  total_obs * p_x
      , chisq = (observed - expected)^2 / expected
    ) %>%
    summarise(
          ll = sum(ll)
        , chisq = sum(chisq)
        , percent_expected = sum(expected > 5) / n()
        , cells = n()
      ) %>%
    mutate(
      BIC = -2 * ll + params * log(total_obs)
      , p.value = pchisq(chisq, df = cells - params -1, lower.tail = FALSE)
      , params = params
    )
  
  if (data2$percent_expected < 0.8) {
    stop("Less than 80% of the cells have more than 5 counts")
  }
  
  return(
    data2 %>%
      select(
        LL = ll
        , `# params` = params
        , BIC = BIC
        , `$\\chi^2\\,p-value$` = p.value
      )
  )
  
}

lcp_results <- 
  lcp_params %>%
  mutate(num = row_number()) %>%
  spread(parameter, estimate) %>%
  select(-num) %>%
  group_by(model) %>%
  summarise(
    lambdas = list(lambda)
    , thetas = list(theta)
  ) %>%
  mutate(
    lambdas = map(lambdas, na.omit)
    , thetas = map(thetas, na.omit)
  ) %>%
  left_join(
    lcp_params %>%
      distinct(model) %>%
      crossing(biz_travel_data) %>%
      nest(N, observed)
    , by = 'model'
  ) %>%
  mutate(pred = pmap(list(data, lambdas, thetas), fn_lcp_ll_results, biz_travel_observed)) %>%
  select(model, pred) %>%
  unnest()
```

Below is a table that summarizes each of the six models fit in this exercise:

```{r final_results, echo = FALSE}
nbd_results %>%
  bind_rows(
    lcp_results
  ) %>%
  pander(caption = "Latent-Class Count Model Comparison")
```

We would select the Zero-Inflated NBD model as our final model. This model has the lowest BIC, only 3 parameters, and a significant $p$-value for the $\chi^2$ goodness-of-fit test. While the 3-segment Poission has a lower log-likelihood, it requires 5 parameters.

# Code

Load business travel dataset

```{r load_data, eval=FALSE}
```

Create functions to find parameters to (zero-inflated) NBD model

```{r NBD Function, eval=FALSE}
```

Find the NBD parameters

```{r nbd_params, eval=FALSE}
```

Create function to find parameters for latent-class poission models

```{r Latent-Class Poission Functions, eval=FALSE}
```

Find the latent-class poission parameters

```{r lcp_params, eval=FALSE}
```

Calculate the NBD results

```{r NBD Results, eval=FALSE}
```

Calculate the latent-class poission results

```{r Poisson Results, eval=FALSE}
```

Output summary of each model

```{r final_results, eval=FALSE}
```

